---
title: "Chase Africa"
subtitle: "Preliminary Analysis"
author: "Maja Zalžnik"
date: "4.12.2018"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: yes
---

***

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H',fig.width=6, fig.height=3.3)
```

```{r echo = FALSE, warnings = FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
df <- readRDS(here::here("data/processed/dandelion.rds"))
df_rounds <- readRDS(here::here( "data/processed/dandelion_rounds.rds"))
df_years <- readRDS(here::here("data/processed/dandelion_years.rds"))


```

# Introduction

This material has been prepared for the introductory meeting between Chase Africa and the Institute of Population Ageing at the start of our collaboration funded by GCRF NGO Secondary Data Funding in the form of a Mini knowledge exchange fellowship. 

It is based on a preliminary analysis of part of the data supplied by Chase Africa and aims to outline and demonstrate some of the possibilities that the data afford. 

*** 
# Data 

Chase Africa has provided six Excel formatted datasets, one for each charity in Kenya or Uganda that they support: 

* `2012-18 CHAT data summary CP.xlsx` -- 7 annual sheets + summary sheet
* `2014-18 Dandelion data summary & CYP CP.xlsx`-- 5 annual sheets + summary sheet
* `2014-18 MKT data summary & CYP CP.xlsx` -- 5 annual sheets + summary sheet
* `Big Life data summary & CYP CP.xlsx` -- 2 annual sheets + summary sheet
* `CHV data summary & CYP CP.xlsx` - 3 annual sheets + summary sheet
* `RICE WN Data summary & CYP CP.xlsx`-- 1 annual sheet + summary sheet

For the purposes of this demonstration only data from *Dandelion* is used for the years 2014-2018 [^1]. The data has been cleaned and consolidated into a single data table. 
The original variables are listed in the appendix, as are the new, derived variables. 

[^1]: All the analysis was performed on all the clinic types i.e. standard and Amboseli as well as Amplify change.  

The data is made *tidy*: 

* each variable forms a single column
* each observation forms a row: a date-venue combination

All derived variables are derived anew, to avoid the possibility of errors inherent in Excel-style cell formulae, and additional variables are derived as well. 

Finally summary statistics are calculated by year and by *funding period* to allow a less granular overview of the trends. 

*** 
# Analysis

## Simple time series

Each of the raw variables on service delivery under the family planning and integrated health services headings can be plotted as a simple time series. Figure \ref{fig:01} shows the simple time series of 5-year implant delivery, both as first time and repeated. 


```{r, echo = FALSE, fig.cap= "\\label{fig:01}Simple time series of 5-year implant delivery (Dandelion: 2014 - 2018)", out.extra = ''}
ggplot(df, aes(date, fp_lt_5yr_1st+fp_lt_5yr_rep)) +
  geom_point() +
  geom_line() +
  # geom_smooth(method = lm, formula = y ~ splines::bs(x, 9),
  #            color = "red") +
  theme_minimal() +
  labs(x = "Date of clinic",
       y = "Number of persons treated")+
  theme(text = element_text(size=8))

```

Figure 2 plots another one of the family planning services provided, this time Depo injections, while Figure 3 shows one of the health services provided: deworming. Again the simple time series plots the number of recipients of each service in each clinic over time. 


```{r, echo = FALSE, fig.cap= "\\label{fig:02}Simple time series of Depo injection delivery (Dandelion: 2014 - 2018)", out.extra = ''}
ggplot(df, aes(date, fp_st_depo_1st+fp_st_depo_rep)) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(x = "Date",
       y = "Number of persons treated")+
  theme(text = element_text(size=8))

```

```{r, echo = FALSE, fig.cap= "\\label{fig:03}Simple time series of deworming", out.extra = ''}
ggplot(df, aes(date, ihs_deworming)) +
  geom_point() +
  geom_line() +
  # geom_smooth(method = lm, formula = y ~ splines::bs(x, 9),
  #            color = "red") +
  theme_minimal() +
  labs(x = "Date",
       y = "Number of persons treated")+
  theme(text = element_text(size=8))

```


Due to the rather high levels of noisiness in the data it makes sense to try and fit a smooth curve to the data, for example using LOESS---locally estimated scatterplot smoothing as in Figure 4 which plots the time series for 3-year implant delivery or Figure 5 which plots the trend in condom uptake. 


```{r, echo = FALSE, fig.cap= "\\label{fig:04}Simple time series of 3-year implant delivery with LOESS curve (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE}
ggplot(df, aes(date, fp_lt_3yr_1st+fp_lt_3yr_rep)) +
  geom_point() +
  geom_line() +
  geom_smooth(span = 0.3, color = "red") +
  theme_minimal() +
  labs(x = "Date",
       y = "Number of persons treated") +
  theme(text = element_text(size=8))

```

```{r, echo = FALSE, fig.cap= "\\label{fig:05}Simple time series of condom delivery with LOESS curve (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE}
ggplot(df, aes(date, fp_condoms)) +
  geom_point() +
  geom_line() +
  geom_smooth(span = 0.3, color = "red") +
  theme_minimal() +
  labs(x = "Date",
       y = "Number of persons treated") +
  theme(text = element_text(size=8))

```

Smoothing should be interpreted carefully though, as it uses a *sliding window* to calculate the best fit, and we can use the size or span of this window to affect how smooth or wiggly we want the curve to be. Figure 6 uses delivery of HIV testing to demonstrate this by plotting a curve with a large span (in blue) and one with a narrower span (red): 


```{r, echo = FALSE, fig.cap= "\\label{fig:06}Simple time series of HIV test delivery with LOESs curve (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE}
ggplot(df, aes(date, ihs_hiv_test)) +
  geom_point() +
  geom_line() +
  geom_smooth(span = 0.2, color = "red", se = FALSE) +
  geom_smooth(span = 0.5, color = "blue", se = FALSE) +
  theme_minimal() +
  labs(x = "Date",
       y = "Number of persons treated")+
  theme(text = element_text(size=8))
```

## Aggregated time series

Due to the nature of the data collection it might be more reasonable to group the clinics together into *funding rounds*, which are groups of 3-15 clinics that are recorded as being funded from a single grant. For example we can compare the total recipients of long term (red) and short term contraceptives (blue) across all the funding periods in Figure \ref{fig:07}.

```{r, echo = FALSE, fig.cap= "\\label{fig:07}Aggregated time series of recipients of long term and short term contraceptives (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE, warning = FALSE}

ggplot(df_rounds, aes(fund_date)) +
  geom_line(aes(y = der_fp_lt_total, colour = "Long term")) + 
  geom_point(aes(y = der_fp_lt_total, colour = "Long term")) + 
  geom_line(aes(y = der_fp_st_total, colour = "Short term")) +
  geom_point(aes(y = der_fp_st_total, colour = "Short term")) +
  scale_color_manual(values = c("red", "blue"), name = "Contraceptive type") +
  theme_minimal() +
  labs(x = "Date of funding round",
       y = "Number of persons treated")+
  theme(text = element_text(size=8))
```


However the problem with this chart, as with the previous simple time series ones, is that clinics and funding periods are not regularly spaced, so the charts don't give a reliable overview of the overall trends. Using the funding rounds makes sense for when we look at costs though, so we will return to them further down. 

But looking at overall trends perhaps the best option is to look at annual changes. Figure \ref{fig:08} summarises the same data as Figure \ref{fig:07} on recipients of long and short term contraceptives, this time looking at the data annually. A quick overview of the individual family planning methods is shown in  Figure \ref{fig:09}. [^2]

```{r, echo = FALSE, fig.cap= "\\label{fig:08}Annualy aggregated time series of recipients of long term and short term contraceptives  (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE, warning = FALSE, fig.height = 3.1}

df_years %>% 
  select(year, der_fp_lt_total, der_fp_st_total) %>% 
  gather(key = var, value = freq, 2:3) %>% 
  ggplot(aes( x = year, y = freq, col = var)) +
  geom_bar(width = 0.4, stat = "identity", position = position_dodge(0.5),fill = "white",
           aes(color = as.factor(var), size = as.factor(year))) +
  scale_color_manual(values = c("red", "blue"), labels = c("Long term", "Short term"), 
                     name = "Contraceptive type") +
  scale_size_manual(values=c(1,1,1,1,.4),guide=F) +
  theme_minimal() +
  labs(x = "Year",
       y = "Number of persons treated")+
  theme(text = element_text(size=8))
```

```{r, echo = FALSE, fig.cap= "\\label{fig:09}Annualy aggregated time series of recipients of all family planning services  (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE, warning = FALSE}
df_years %>% 
  select(year, starts_with("fp")) %>% 
  mutate(fp_lt_5yr = fp_lt_5yr_1st + fp_lt_5yr_rep,
         fp_lt_3yr = fp_lt_3yr_1st + fp_lt_3yr_rep,
         fp_st_depo = fp_st_depo_1st + fp_st_depo_rep) %>% 
  select(-fp_condoms, -fp_lt_5yr_1st, -fp_lt_5yr_rep, 
         -fp_lt_3yr_1st, -fp_lt_3yr_rep, 
         -fp_st_depo_1st, -fp_st_depo_rep) %>% 
  gather(key = var, value = freq, 2:11) %>% 
  filter(!is.na(freq))  %>% 
  mutate(var = factor(var, labels = c("3 year implant", "5 year implant", "IUCD",
                                        "Depo injection", "Pills 1 month", "Pills 3 months"))) %>% 
  ggplot(aes(x = year, y = freq)) +
  geom_bar(width = 0.4, stat = "identity", position = position_dodge(0.5), color = "red", fill = "white",
           aes( size = as.factor(year))) +
  facet_wrap("var") +
  scale_size_manual(values=c(1,1,1,1,.4),guide=F) +
  theme_minimal() +
  labs(x = "Year",
       y = "Number of persons treated")+
  theme(text = element_text(size=8))
``` 

[^2]:  Because the 2018 data is still incomplete, last two bars are drawn thinner to make that clear.  

## Couple Years of Protection Analysis

Using the amounts of funding for each funding period and the standard conversion factors for each type of contraceptive we can calculate the cost of *Couple years of protection* (in GBP) and see how it has varied over the funding rounds in Figure \ref{fig:10}.


```{r, echo = FALSE, fig.cap= "\\label{fig:10}Cost (in GBP) of couple year of protection provided (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE, warning = FALSE}
df_rounds %>% 
  ggplot(aes(fund_date, der_gpb_per_cyp)) +
  geom_point() +
  geom_line() +
    theme_minimal() +
  labs(x = "Date of funding round",
       y = "GBP per CYP")+
  theme(text = element_text(size=8))

```

Using CYP we can also investigate how the shares of different type of contraceptives have changed over time in terms of how many CYPs they provide: 

```{r, echo = FALSE, fig.cap= "\\label{fig:11}Contribution to total CYP provided by each contraceptive type (Dandelion: 2014 - 2018)", out.extra = '', message = FALSE, warning = FALSE}

df_rounds %>% 
  select(fund_date,
         der_fp_lt_iucd_cyp,
         der_fp_implants_tot_cyp,
         der_fp_depo_tot_cyp,
        der_fp_pills_tot_cyp) %>% 
  gather(key = contr_type, value = cyp, 2:5) %>% 
  group_by(fund_date) %>% 
  mutate(prop.cyp = cyp/sum(cyp)) %>% 
  arrange(fund_date) %>% 
  ggplot(aes(x = fund_date, y = prop.cyp)) +
  geom_line(aes(color = contr_type)) + 
  geom_point(aes(color = contr_type)) + 
scale_color_manual(values = c("red", "blue", "goldenrod2", "forestgreen"),
                   labels = c("Depo injections", "Implants", "IUCDs", "Pills"),
                           name = "Contraceptive type") +
  theme_minimal() +
  labs( x = "Date of funding round",
       y = "Proportion of total CYP provided")+
  theme(text = element_text(size=8))
  
  
```
\newpage
# Appendix {-}

The variables fall into the following categories:

* `id` variables such as `date`, and `venue`
* `fund_` variables have been extracted from the single column in the excel files relating to the funding round, amounts and dates.
* `fp_` variables are to do with family planning provision and are further split into `fp_lt_` and `fp_st_` for long and short term contraception. 
* `ihs_` variables refer to integrated health care services
* `der_` are variables derived from the basic variables listed above (if they had already been calculated in the original excel files, they are still re-calculated). 

```{r echo = FALSE}
library(kableExtra)
var_list <- readRDS(here::here("data/processed/var_list.rds"))
knitr::kable(var_list, "latex", longtable = TRUE, booktabs = TRUE) %>% 
  kable_styling() %>% 
  group_rows("ID", 1,2) %>% 
  group_rows("Family Planning - Long Term", 3,7) %>% 
  group_rows("Family Planning - Short Term", 8,13) %>% 
  group_rows("Family Planning - Ohter", 14,18) %>% 
  group_rows("Integrated Health Care", 19,30) %>% 
  group_rows("Funding Information", 31,35) %>% 
  group_rows("Numbers of recipients", 36,40) %>% 
  group_rows("Couple Years of Protection", 41,57) %>% 
  group_rows("Costs", 58,61)

```