---
title: "Chase Africa"
subtitle: "Preliminary Analysis"
author: "Maja Zalžnik"
date: "4.12.2018"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

```{r echo = FALSE, warnings = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
df <- readRDS(here::here("data/processed/dandelion.rds"))
df_rounds <- readRDS(here::here( "data/processed/dandelion_rounds.rds"))
df_years <- readRDS(here::here("data/processed/dandelion_years.rds"))


```

# Introduction

This material has been prepared for the introductory meeting between Chase Africa and the Institute of Population Ageing at the start of our collaboration funded by GCRF NGO Secondary Data Funding in the form of a Mini knowledge exchange fellowship. 

It is based on a preliminary analysis of part of the data supplied by Chase Africa and aims to outline and demonstrate some of the possibilities that the data afford. 

# Data 

Chase Africa has provided six Excel formatted datasets, one for each charity in Kenya or Uganda that they support: 

* `2012-18 CHAT data summary CP.xlsx` -- 7 annual sheets + summary sheet
* `2014-18 Dandelion data summary & CYP CP.xlsx`-- 5 annual sheets + summary sheet
* `2014-18 MKT data summary & CYP CP.xlsx` -- 5 annual sheets + summary sheet
* `Big Life data summary & CYP CP.xlsx` -- 2 annual sheets + summary sheet
* `CHV data summary & CYP CP.xlsx` - 3 annual sheets + summary sheet
* `RICE WN Data summary & CYP CP.xlsx`-- 1 annual sheet + summary sheet

For the purposes of this demonstration only data from *Dandelion* is used for the years 2014-2018 [^1]. The data has been cleaned and consolidated into a single data table. 
The original variables are listed in the appendix, as are the new, derived variables. 

[^1]: All the analysis was performed on all the clinic types i.e. standard and Amboseli as well as Amplify change.  

The data is made *tidy*: 

* each variable forms a single column
* each observation forms a row: a date-venue combination

All derived variables are derived anew, to avoid the possibility of errors inherent in Excel-style cell formulae, and additional variables are derived as well. 

Finally summary statistics are calculated by year and by *funding period* to allow a less granular overview of the trends. 

# Analysis

## Simple time series

Each of the raw variables on service delivery under the family planning and integrated health services headings can be plotted as a simple time series. 

```{r, echo = FALSE, fig.cap= "Simple time series of implant delivery", out.extra = '',fig.height = 4, fig.width = 6}
ggplot(df, aes(date, fp_lt_5yr_1st+fp_lt_5yr_rep)) +
  geom_point() +
  geom_line() +
  # geom_smooth(method = lm, formula = y ~ splines::bs(x, 9),
  #            color = "red") +
  theme_minimal() +
  labs(title = "5 year implants (Dandelion: 2014 - 2018)",
       x = "Date",
       y = "Number of persons treated")

```

```{r, echo = FALSE, fig.cap= "Simple time series of deworming", out.extra = '', fig.height = 4, fig.width = 6}
ggplot(df, aes(date, ihs_deworming)) +
  geom_point() +
  geom_line() +
  # geom_smooth(method = lm, formula = y ~ splines::bs(x, 9),
  #            color = "red") +
  theme_minimal() +
  labs(title = "Deworming (Dandelion: 2014 - 2018)",
       x = "Date",
       y = "Number of persons treated")

```


Due to the noisiness of the data it makes sense to fit a smooth curve to the data, for example: 


```{r, echo = FALSE, fig.cap= "Simple time series of 3-year implant delivery", out.extra = '',fig.height = 4, fig.width = 6}
ggplot(df, aes(date, fp_lt_3yr_1st+fp_lt_3yr_rep)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 9),
              color = "red") +
  theme_minimal() +
  labs(title = "3 year implants (Dandelion: 2014 - 2018)",
       x = "Date",
       y = "Number of persons treated")

```



```{r, echo = FALSE, fig.cap= "Simple time series of HIV test delivery", out.extra = '',fig.height = 4, fig.width = 6}
ggplot(df, aes(date, ihs_hiv_test)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 9),
              color = "red") +
  theme_minimal() +
  labs(title = "HIV tests conducted (Dandelion: 2014 - 2018)",
       x = "Date",
       y = "Number of persons treated")

```
# Appendix {-}

The variables fall into the following categories:

* `id` variables such as `date`, and `venue`
* `fund_` variables have been extracted from the single column in the excel files relating to the funding round, amounts and dates.
* `fp_` variables are to do with family planning provision and are further split into `fp_lt_` and `fp_st_` for long and short term contraception. 
* `ihs_` variables refer to integrated health care services
* `der_` are variables derived from the basic variables listed above (if they had already been calculated in the original excel files, they are still re-calculated). 

```{r echo = FALSE}
library(kableExtra)
var_list <- readRDS(here::here("data/processed/var_list.rds"))
knitr::kable(var_list, "latex", longtable = TRUE, booktabs = TRUE) %>% 
  kable_styling() %>% 
  group_rows("ID", 1,2) %>% 
  group_rows("Family Planning - Long Term", 3,7) %>% 
  group_rows("Family Planning - Short Term", 8,13) %>% 
  group_rows("Family Planning - Ohter", 14,18) %>% 
  group_rows("Integrated Health Care", 19,30) %>% 
  group_rows("Funding Information", 31,35) %>% 
  group_rows("Numbers of recipients", 36,40) %>% 
  group_rows("Couple Years of Protection", 41,57) %>% 
  group_rows("Costs", 58,61)

```